\documentclass[preprint,11pt]{article}
%\documentclass{elsart3-1}

\usepackage{amsthm,amsmath,amssymb,amsfonts,amscd,amsbsy}%,latexsym,epsfig,subfig}
%\usepackage{graphicx,psfrag,verbatim,wasysym}
\usepackage{graphicx}
\usepackage[ruled,noline,linesnumbered]{algorithm2e}
%\usepackage{bm}
\usepackage[tight]{subfigure}
\usepackage{fullpage}
\usepackage{color}
%\usepackage{setspace}
%\usepackage[tight]{subfigure}
%\usepackage{listings}
\usepackage{enumerate}
\newcommand{\doi}[1]{\textsc{doi:} \textsf{#1}\xspace}

\DeclareGraphicsExtensions{.jpg,.png,.pdf}
\DeclareGraphicsRule{*}{mps}{*}{}

%\setlength{\parindent}{0em}
\setlength{\parskip}{0em}

\include{commands}

%\renewcommand{\theenumi}{(\alph{enumi})}
% 
%
%

\begin{document}

\noindent\begin{Large}\textbf{AER1418 Problem Set 1}\end{Large} \hfill Due Tuesday, 6 February 2018, at 14:00.

\vspace{1em}
\noindent A hard copy of the problem set is due in class at the specified time.  %For problems that require coding, please turn in a hard copy of the program listings and also upload the code to the Portal to facilitate the grading process.
Please adhere to the collaboration policy: the final write up must be prepared individually without consulting others. (See the syllabus for details.)

\section*{Part 1. Formulation and analysis}
We wish to understand the insulation characteristics of a jacket worn by a person walking outside in a cold, windy day.  In our simplified analysis, the person is a circle that acts as a heat source, and the jacket is a ring that surrounds the person. We wish to model the temperature distribution in the ring-shaped jacket and to evaluate the temperature on the inner surface, which contacts the person.

We now provide a mathematical description of the problem.  All variables are nondimensionalized.  We introduce a two-dimensional annular domain
\begin{equation*}
  \Omega_{\rm annulus} \equiv \{ x \in \RR^2 \ | \ r_{\rm in} < \sqrt{x_1^2 + x_2^2} < r_{\rm out} \},
\end{equation*}
where $r_{\rm in}$ and $r_{\rm out}$ are the inner and outer radius, respectively, so that $0 < r_{\rm in} < r_{\rm out}$. The associated inner and outer boundaries are
\begin{align*}
  \Gamma_{\rm in} &\equiv \{ x \in \RR^2 \ | \ \sqrt{x_1^2 + x_2^2} = r_{\rm in} \}, \\
  \Gamma_{\rm out} &\equiv \{ x \in \RR^2 \ | \ \sqrt{x_1^2 + x_2^2} = r_{\rm out} \}.
\end{align*}
The steady-state temperature distribution inside the annulus is modeled by the (steady-state) heat equation
\begin{equation*}
  - \nabla \cdot (k \nabla u) = 0 \quad \text{in } \Omega_{\rm annulus} ,
\end{equation*}
where $k > 0$ is the thermal conductivity. The heat transfer at the inner wall $\Gamma_{\rm in}$ from the heat source is modeled by
\begin{equation*}
  - n \cdot (k \nabla u) = -g \quad \text{on } \Gamma_{\rm in},
\end{equation*}
where $n$ is the outward-pointing normal with respect to $\Omega$, and $g > 0$ is the heat flux. %; here, the negative flux indicates that the energy is transferred from the heat source to the annulus.
The heat transfer at the outer wall $\Gamma_{\rm out}$ to the air is modeled by 
\begin{equation*}
  - n \cdot (k \nabla u) = B_{\rm out} (u - u_{\infty})  \quad \text{on } \Gamma_{\rm out},
\end{equation*}
where $B_{\rm out} > 0$ is the Biot number that characterizes the heat transfer to the surrounding air.  Without loss of generality, we set $u_{\infty} = 0$; i.e., the temperature $u$ is defined with respect to the ambient air temperature.

Answer the following questions:
\begin{enumerate}[(a)]
\item Derive a variational form of the two-dimensional heat conduction problem.  Specifically, identify (i) the trial and test spaces, (ii) bilinear form, and (iii) linear form.
\item The two-dimensional problem is axisymmetric. Reformulate the variational problem as a one-dimensional variational problem on the domain $\Omega \equiv (r_{\rm in}, r_{\rm out}) \subset \RR^1$.  Specifically, identify (i) the trial and test spaces, (ii) bilinear form, and (iii) linear form.

  \emph{Note:} the integrand of the bilinear form should depend on the coordinate $x \in \Omega \subset \RR^1$.
\end{enumerate}
For the following questions, the variational problem, bilinear form, and linear form refer to those obtained in (b).
\begin{enumerate}[(a)]
  \setcounter{enumi}{2}
\item Show that the bilinear form is coercive.
\item Show that the bilinear form is continuous.
\item Show that the linear form is continuous.
\item Prove that the variational problem has a unique solution.
\item Suppose the Biot number is zero: $B_{\rm out} = 0$.  Prove that the bilinear form is not coercive.  Does the problem have a unique solution?

  \emph{Note:} to prove that the bilinear from is not coercive, you need to find only one function for which the coercivity condition is violated.
\item State a minimization form of the variational problem.  Specifically, identify (i) the function space and (ii) energy functional.
\end{enumerate}

\section*{Part 2. Implementation}
We now wish to develop a finite-element code that approximates the solution to the one-dimensional variational problem obtained in Part 1. The code should use a piecewise linear finite element space and accept an arbitrary discretization parameter $h > 0$.  
\begin{enumerate}[(a)]
\item Write a finite-element code that solves the heat conduction problem.

  \emph{Note.} We need to evaluate the entries of the stiffness matrix $\hat A_h$ and load vector $\hat f_h$.  For example, to evaluate the diagonal entries of $\hat A_h$, consider a hat-shaped piecewise-linear basis function delineated by points $x_{i-1}$, $x_i$, and $x_{i+1}$ with the support $(x_{i-1},x_{i+1})$ and the peak at $x_i$; then, express the diagonal entry as a function of the coordinates $x_{i-1}$, $x_{i}$, and $x_{i+1}$.  Evaluate all other entries of the stiffness matrix and load vector in a similar manner. 
%\item Evaluate all other entries of the stiffness matrix.  Introduce an appropriate set of points that delineates basis functions, and express entries as the coordinates of those points.
%\item Write a finite-element code that solves the heat conduction problem.
\item Plot the finite-element solution $u_h$ for $r_{\rm in} = 1/2$, $r_{\rm out} = 2$, $k = 1$, $g = 1$, $B_{\rm out} = 1$, and a uniform mesh spacing of $h = 1/4$.  
\item Consider the output $\ell^{o}(u) \equiv g r_{\rm in} u(x = r_{\rm in})$. Show that
  \begin{equation*}
    \ell^o(u) - \ell^o(u_h) = a(e,e),
  \end{equation*}
  where $e \equiv u - u_h$ is the error in the finite element approximation $u_h$.  
\item Solve the finite element problem for $r_{\rm in} = 1/2$, $r_{\rm out} = 2$, $k = 1$, $g = 1$, $B_{\rm out} = 1$, and uniform grid spacings of $h = 1/2, 1/4, 1/8, \dots, 1/64$. Report the reference value of the output for $h_{\rm ref} \equiv 1/512$, $\ell^o(u_{\rm ref})$.  Plot the output error (with respect to $\ell^o(u_{h_{\rm ref}})$) against the grid spacing $h$ in log-log scale, and report the observed convergence rate.  Does the observed convergence rate match your expectation based on the theory?
%\item Execute the finite-element code for $B_{\rm out} = 0$.  %How does the code fail?
%  Does the observed behavior match your expectation?
\end{enumerate}


\end{document}

